# Data

## Description

Our dataset was meticulously constructed through a multi-step process combining data from the International Monetary Fund's (IMF) DataMapper platform (https://www.imf.org/external/datamapper/datasets) and the United Nations classification databases. The economic indicators were initially downloaded as separate Excel files from the IMF, including GDP Growth, GDP Per Capita, Unemployment rates, Inflation rates, Population statistics, Current Account balances, and Government Debt levels, all covering the period from 2018 to 2023. The data processing began with standardizing country names across all files and removing regional aggregates to focus on country-level analysis. We then used VLOOKUP functions in Excel to merge data from different files and combined all separate sheets into a single Excel workbook. A crucial step involved manually correcting the development status classifications due to inconsistencies in country naming conventions between IMF and UN datasets. Throughout this process, we encountered and addressed several challenges, including inconsistent country naming conventions, missing data points (particularly for developing nations), and varying data availability across different time periods and indicators. Rather than imputing values, we preserved original "no data" entries to maintain data integrity. The final dataset encompasses information for 196 countries, though coverage varies by indicator and year, providing a comprehensive yet transparent view of global economic indicators that acknowledges data limitations and gaps. The IMF updates this data semi-annually, while the UN classifications are updated annually, ensuring our analysis reflects recent global economic trends while maintaining consistent categorical classifications.

### Data Sources

Our analysis combines data from two authoritative sources:

1.  **International Monetary Fund (IMF) DataMapper**: We've collected extensive economic indicators from 2018 to 2023 including:
    -   GDP Growth rates
    -   GDP Per Capita figures
    -   Unemployment rates
    -   Inflation rates
    -   Population statistics
    -   Current Account balances
    -   Government Debt levels
2.  **United Nations Classification**: We've incorporated:
    -   Regional classifications
    -   Development status categorizations (Developed vs. Developing nations)

## Missing value analysis

```{r, MESSAGE = FALSE, WARNING = FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(tibble)
library(reshape2)
```

```{r}
data <- read.csv("Merged_data.csv")[,-1]
```

```{r, fig.height = 6}
missing_percentage <- colSums(is.na(data)) / nrow(data) * 100
missing_percentage <- missing_percentage[missing_percentage > 0]
missing_df <- data.frame(Column = names(missing_percentage), Percentage = missing_percentage)

ggplot(missing_df, aes(x = reorder(Column, -Percentage), y = Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Percentage of Missing Values by Column", x = "Columns", y = "Percentage Missing (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1), plot.title = element_text(hjust = 0.5))
```

The first graph illustrates the percentage of missing data for each column across the entire dataset. Variables related to unemployment (Unemployment_2018, Unemployment_2019, etc.) stand out as having a high proportion of missing data, with over 40% of their values missing. This suggests that data collection or reporting issues for unemployment metrics may be widespread. Other variables, such as Government_Debt and Current_Account across various years, have significantly lower levels of missingness, with most showing only minor gaps.


```{r}
missing_stacked <- data |>
  group_by(Region) |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(-Region, 
               names_to = "Variable", 
               values_to = "Percentage") |>
  filter(Percentage > 0) |>  # Only keep variables with missing values
  mutate(Variable = gsub("_[0-9]{4}$", "", Variable)) |>
  group_by(Variable, Region) |>
  summarise(Percentage = mean(Percentage), .groups = "drop") |>
  filter(!is.na(Region))


ggplot(missing_stacked, 
       aes(x = reorder(Variable, Percentage), 
           y = Percentage, 
           fill = Region)) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Missing Values Distribution by Variable and Region",
       x = "Variables",
       y = "Percentage Missing (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),
    axis.text.y = element_text(size = 8),
    legend.position = "right",
    legend.text = element_text(size = 8))
```
The second graph highlights the regional distribution of missing values for key variables. Unemployment data shows consistent missingness across all regions, indicating global reporting challenges. In contrast, Government_Debt data has significant gaps in specific regions, such as Africa and the Middle East, while other regions are more complete. Current_Account missingness is relatively minor and localized to a few regions. 
