# Results

```{r, MESSAGE=FALSE, WARNING=FALSE}
library(ggplot2)
library(dplyr)
library(viridis)
library(tidyr)
library(scales)
library(viridis)
library(ggridges)
library(corrplot)
library(factoextra)
library(tibble)
library(gridExtra)
library(treemapify)
library(FactoMineR) 
library(factoextra)
library(parcoords)
library(reshape2)
```

```{r}
data <- read.csv("Merged_data.csv")[,-1]
```

## Graph 1: Average GDP Growth by Region and Development Status (2023)
```{r}
# Calculate average GDP growth
average_gdp_growth <- data |>
  mutate(GDP_Growth_2023 = as.numeric(ifelse(GDP_Growth_2023 == "no data", NA, GDP_Growth_2023))) |>
  group_by(Region, Status) |>
  summarize(
    Average_GDP_Growth = mean(GDP_Growth_2023, na.rm = TRUE),
    .groups = "drop")

ggplot(average_gdp_growth, 
       aes(x = reorder(Region, Average_GDP_Growth), 
           y = Average_GDP_Growth, 
           fill = Status)) +
  geom_bar(stat = "identity", 
           position = "stack",
           width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", Average_GDP_Growth)),
            position = position_stack(vjust = 0.5),
            color = "white",
            size = 3.5) +
  labs(title = "Average GDP Growth by Region and Development Status (2023)",
    subtitle = "Regions ordered by total GDP growth",
    x = NULL,
    y = "Average GDP Growth Rate (%)",
    fill = "Development Status") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5),
    axis.title.y = element_text(size = 12, margin = margin(r = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank())
```
This chart illustrates the average GDP growth rates across different regions in 2023, breaking the data down by development status. Asia leads the pack with the highest overall growth at 18.3%, driven largely by developed economies at 15.0%, while developing economies contributed 3.3%. In contrast, the Middle East experienced negative growth at -5.4%, entirely from its developing economies.

Europe showed modest growth at 3.1%, with developed economies contributing 1.4% and developing economies slightly higher at 1.7%. Oceania performed slightly better, with a total growth of 4%, split between 1.3% from developed economies and 2.7% from developing economies. North America recorded 5.2% growth overall, with 2.0% coming from developed economies and 3.2% from developing ones.

Africa and South America, made up entirely of developing economies, reported steady growth rates of 3.7% and 4.3%, respectively. Meanwhile, the "Other" category also performed well, growing by 5.3%, solely from developing economies. The data highlights some interesting patterns. While many regions rely heavily on either developed or developing economies, others like Asia show a mix, with its strong overall growth largely driven by its developed economies. This is notable, as developing economies are typically the ones with higher growth rates.

## Graph 2: GDP Growth Trends by Region and Development Status (2018-2023)
```{r}
# Transform data into long format
gdp_long <- data |>
  pivot_longer(cols = starts_with("GDP_Growth"),
    names_to = "Year",
    values_to = "GDP_Growth") |>
  mutate(Year = as.numeric(gsub("GDP_Growth_", "", Year)),
    GDP_Growth = as.numeric(ifelse(GDP_Growth == "no data", NA, GDP_Growth))) |>
  filter(!is.na(GDP_Growth)) 

# Plot GDP growth over time
ggplot(gdp_long, aes(x = Year, y = GDP_Growth, color = Status, group = interaction(Status, Region))) +
  geom_line(size = 0.8) +
  geom_point(size = 1.5) +
  facet_wrap(~ Region, scales = "free", ncol = 3) +
  labs(title = "GDP Growth Trends by Region and Development Status (2018-2023)",
    subtitle = "GDP growth trends over time for developed and developing nations",
    x = "Year",
    y = "GDP Growth Rate (%)",
    color = "Development Status") +
  scale_color_manual(values = c("Developed" = "#1f77b4", "Developing" = "#ff7f0e")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 11, face = "bold"))

```
In 2023, Asia stood out as the global growth leader, achieving an impressive 18.3% total GDP growth rate, while the Middle East struggled with a notable decline, reporting -5.4% growth. This stark contrast between regions led us to a deeper dive into the historical trends through time series analysis.

The time series analysis revealed the severe economic impact of COVID-19 in 2020, which caused significant contractions across most regions. Europe, the Middle East, and the "Other" category were among the hardest hit during this period. However, recovery patterns differed widely. Asia showcased exceptional resilience, with both its developed and developing economies maintaining stable growth through the post-pandemic years. Europe, while rebounding strongly in 2021, saw its growth stabilize at a lower rate in subsequent years. North America displayed a slower, steadier recovery, with both developed and developing economies following a gradual upward trajectory.

The analysis also highlighted clear differences between developed and developing economies. Developed economies (depicted by blue lines) generally experienced lower but more stable growth rates. In contrast, developing economies (shown in orange lines) were characterized by higher volatility but also greater growth potential. This distinction was particularly pronounced in South America's performance, where developing economies demonstrated notable growth following the pandemic.

## Graph 3: Regional Government_Debt Distribution (2018–2023)
```{r, fig.width = 8}
library(ggridges)
data |>
  select(starts_with("Government_Debt"), Region) |>
  pivot_longer(-Region, names_to = "Year", values_to = "Government_Debt") |>
  mutate(Year = gsub("Government_Debt_", "", Year)) |>
  ggplot(aes(x = as.numeric(Government_Debt), y = Region, fill = Region)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5) +
  facet_wrap(~Year) +
  labs(title = "Regional Government Debt Distribution (2018–2023)",
       x = "Government Debt (% of GDP)", y = "Region") + # Updated label
  theme_ridges() +
  theme(legend.position = "none")
```
Building on our earlier analysis of GDP growth trends, we now turn to examining government debt patterns from 2018 to 2023. The ridge density plots provide valuable insights into how debt distributions have evolved across regions over time.

Europe consistently exhibits the widest spread in government debt levels, with a distinct bimodal distribution suggesting two clusters of countries with differing debt strategies. This pattern remains largely stable throughout the period, although a marked increase in overall debt levels is evident post-2020, likely due to pandemic-related fiscal measures. South America shows high variability in government debt levels, with a wide and increasingly dispersed distribution after 2020. The long right tails indicate a few countries with exceptionally high debt levels, reflecting significant outliers in the region's fiscal landscape. 

Asia displays a more concentrated debt distribution, with most countries falling within a moderate debt range. However, some outliers with very high debt levels are evident, as highlighted by the right tails in the distributions, signaling unique cases of heavy borrowing. Africa demonstrates relatively lower average debt levels compared to other regions, with a fairly consistent distribution throughout the period. After 2020, however, the distribution begins to widen slightly, suggesting growing diversity in debt levels across African nations.

The Middle East has one of the most compact debt distributions, with little year-to-year variability. This consistency indicates that the government has tighter control over its debt than other regions. A notable trend across all regions is a clear shift in debt distributions around 2020, which corresponds to the global economic impact of the COVID-19 pandemic. Most regions experienced an increase in debt levels and distributions following this period, reflecting widespread increases in government borrowing to address the crisis. We can describe the various approaches and challenges that regions face when managing government debt in the aftermath of global economic disruptions.

## Graph 4: Log-Scaled GDP Growth vs. Per Capita Income Across Regions (2022 vs 2023)

This shows regional variations in GDP growth against per capita income in 2023. Though their GDP per capita is usually low, African nations show great variation in growth rates, which reflect different phases of development. Typical of advanced economies, European nations are concentrated at higher income levels with steady, modest growth. Asia exhibits the most varied trends; large-population countries are showing strong development all over a wide income range. While South America groups more consistently in the middle for both measures, North America sits in the higher income range with consistent increase. With some high-income countries and differing rates of development, Oceania and the Middle East show erratic trends. While smaller countries are distributed over all income levels, larger bubbles reflect populous countries like those in Asia, which mix great growth and mid-to-high income.

```{r, fig.height =4, fig.width =12}
gdp_income_data_long <- data |>
  select(Country, Region, Status, 
         starts_with("Population"), 
         starts_with("GDP_Growth"), 
         starts_with("GDP_Per_Capi")) |>
  pivot_longer(cols = starts_with("GDP_Growth"), 
               names_to = "Year", 
               values_to = "GDP_Growth") |>
  mutate(Year = gsub("GDP_Growth_", "", Year)) |>
  (\(df) df |>
    pivot_longer(cols = starts_with("GDP_Per_Capi"), 
                 names_to = "Year_Per_Capi", 
                 values_to = "GDP_Per_Capi") |>
    filter(Year == gsub("GDP_Per_Capi_", "", Year_Per_Capi)))() |>
  (\(df) df |>
    pivot_longer(cols = starts_with("Population"), 
                 names_to = "Year_Population", 
                 values_to = "Population") |>
    filter(Year == gsub("Population_", "", Year_Population)))() |>
  select(-Year_Per_Capi, -Year_Population) |>
  mutate(across(c(GDP_Growth, GDP_Per_Capi, Population), 
                ~as.numeric(ifelse(. == "no data", NA, .)))) |>
  na.omit()

gdp_income_data_filtered <- gdp_income_data_long |>
  filter(Year %in% c("2022", "2023"))

plot_list <- split(gdp_income_data_filtered, gdp_income_data_filtered$Year) |>
  lapply(function(year_data) {
    ggplot(year_data, aes(x = GDP_Growth, 
                          y = GDP_Per_Capi, 
                          color = Region, 
                          size = Population)) +
      geom_point(alpha = 0.7) +
      scale_size(range = c(2, 10), name = "Population") +
      scale_color_brewer(palette = "Set1") +
      scale_x_log10(name = "Log of GDP Growth Rate (%)") +
      scale_y_log10(name = "Log of GDP Per Capita (USD)",
                    labels = label_number(scale = 0.001, suffix = "k")) +
      labs(
        title = paste("Year:", unique(year_data$Year)),
        color = "Region"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 12),
        legend.position = "right",
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10)
      )
  })

combined_plot <- do.call(grid.arrange, c(plot_list, ncol = 2))
```
There are some intriguing changes when comparing the performance of nations in 2022 and 2023.

In 2022, things were fairly predictable: European countries were comfortably at the top, with high living standards (high GDP per capita), but not growing at a rapid pace. African countries experienced varying degrees of growth while maintaining lower overall living standards. Pretty much what you would expect.There are some interesting changes when we move to 2023. Growth rates have become more widely distributed overall, with some nations seeing much slower growth and others seeing much higher growth rates (up to almost 100% on the log scale). Although their growth rates varied more than in 2022, African nations continued to fall into the lower GDP per capita range.

Over the course of both years, Asian nations (in blue), especially those with larger populations (shown by larger circles), remained largely stable, with some exhibiting better growth rates in 2023. The GDP per capita of European nations remained high in 2023, albeit at somewhat more cautious growth rates than in 2022. The comparison of these years indicates that, although basic regional GDP per capita trends stayed mostly stable, 2023 saw greater economic volatility and greater differences in growth rates. The continuous global economic recovery and disparate regional reactions to post-pandemic issues are probably to blame for this.


## Graph 5: Relationship Between Government Debt and Economic Growth (2023)
```{r, fid.width = 10, fig.height = 6}
debt_growth_data <- data |>
  mutate(
    GDP_Growth_2023 = as.numeric(ifelse(GDP_Growth_2023 == "no data", NA, GDP_Growth_2023)),
    Government_Debt_2023 = as.numeric(ifelse(Government_Debt_2023 == "no data", NA, Government_Debt_2023))
  ) |>
  (\(df) df[!is.na(df$GDP_Growth_2023) & !is.na(df$Government_Debt_2023), ])()


avg_metrics <- debt_growth_data |>
  mutate(Growth_Bin = cut(GDP_Growth_2023, 
                         breaks = seq(-10, 20, by = 2),
                         labels = paste(seq(-10, 18, by = 2), 
                                      seq(-8, 20, by = 2), 
                                      sep = "-"))) |>
  group_by(Growth_Bin) |>
  summarise(
    Avg_Debt = mean(Government_Debt_2023, na.rm = TRUE),
    Count = n())

ggplot() +
  geom_col(data = avg_metrics,
           aes(x = Growth_Bin, y = Count, fill = Avg_Debt),
           alpha = 0.8) +
  geom_point(data = debt_growth_data,
             aes(x = cut(GDP_Growth_2023, 
                        breaks = seq(-10, 20, by = 2),
                        labels = paste(seq(-10, 18, by = 2), 
                                     seq(-8, 20, by = 2), 
                                     sep = "-")),
                 y = Government_Debt_2023,
                 color = Region),
             position = position_jitter(width = 0.2),
             alpha = 0.7) +
  scale_fill_viridis(name = "Average\nGovernment Debt (%)",
                     option = "plasma") +
  scale_y_continuous(
    name = "Government Debt (% of GDP)",
    sec.axis = sec_axis(~./5, name = "Number of Countries")) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Government Debt vs Economic Growth (2023)",
    subtitle = "Bars show # of countries in each growth range, colored by average debt level",
    x = "GDP Growth Rate Range (%)",
    color = "Region") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.grid.minor = element_blank())
```
This visualization explores the relationship between government debt and economic growth across regions in 2023, revealing nuanced patterns. Most countries cluster within the 0-4% growth range, particularly 2-4%, with average debt levels around 60-70% of GDP (indicated by pink/purple bars). High debt levels (over 150% of GDP) appear across various growth ranges, showing that high debt does not always equate to low growth; for example, some Asian countries sustain strong growth with substantial debt, while certain European nations maintain stable growth despite high debt levels. Lower-growth regions (-2-0%) tend to have slightly higher average debt, but this trend is not uniform—African nations often achieve higher growth with lower debt, whereas European countries display moderate growth alongside higher debt burdens. Outliers in extreme growth ranges (below -6% or above 8%) often reflect unique economic conditions rather than typical debt-growth relationships. While there is a mild negative association between debt and growth, the relationship is complex and shaped by regional context and economic structure, challenging the assumption that high debt universally inhibits growth.

## Graph 6: Government Debt Trends by Region (2018–2023)
```{r}
data |>
  select(starts_with("Government_Debt"), Region) |>
  pivot_longer(-Region, names_to = "Year", values_to = "Government_Debt") |>
  mutate(Year = as.numeric(gsub("Government_Debt_", "", Year))) |>
  group_by(Year, Region) |>
  summarise(Average_Debt = mean(as.numeric(Government_Debt), na.rm = TRUE)) |>
  ggplot(aes(x = Year, y = Average_Debt, color = Region)) +
  geom_line(size = 1.2) +
  geom_point() +
  geom_text(aes(label = ifelse(Average_Debt > 70, paste0(round(Average_Debt, 1), "%"), "")),
            hjust = 0.5, vjust = -0.5) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) + # Format y-axis as percentage
  labs(title = "Government Debt Trends by Region (2018–2023)",
       x = "Year", y = "Average Debt (% of GDP)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10))
```
After analyzing trends in GDP growth, let's look at how government debt changed geographically between 2018 and 2023. This tells an intriguing tale, particularly in relation to the COVID-19 era.

The most dramatic increase occurs in 2020, when the "Other" region reaches a high of 112.3 percent of GDP. South America followed a similar pattern, increasing to 92.2%, while North America reached approximately 80%. This surge in debt levels demonstrates how governments around the world responded to the pandemic with massive spending. The recovery pattern since then has been particularly interesting. Most regions show a gradual reduction in debt levels after 2020, indicating efforts to regain financial control. The "Other" region, for example, has seen its debt fall significantly from its peak of 112.3% to 78.7% by 2023.

Throughout this time, the debt levels of Europe and Asia have remained comparatively constant, averaging 60%. At roughly 50% of GDP, the Middle East has the lowest and most stable debt profile of any region. The most conservative approach to debt management among all regions is demonstrated by Oceania, which stands out for maintaining remarkably stable debt levels even during the 2020 crisis.

Some of the patterns of GDP growth that we have seen are put into perspective by these trends. South America and other regions that took on more debt during the pandemic appear to be having more difficulty maintaining growth. Asia and Europe, on the other hand, are exhibiting more consistent economic performance due to their more stable debt levels.

## Graph 7: Correlation of Government Debt and GDP Growth
```{r}
# Select relevant columns for government debt and GDP growth
relevant_data <- data[, grepl("Government_Debt|GDP_Growth", colnames(data))]

# Remove rows with missing values
relevant_data <- na.omit(relevant_data)

# Calculate the correlation matrix
cor_matrix <- cor(relevant_data, use = "pairwise.complete.obs")

# Subset the correlation matrix: Debt on Y-axis, GDP Growth on X-axis
debt_vars <- grep("Government_Debt", colnames(cor_matrix), value = TRUE)
gdp_vars <- grep("GDP_Growth", colnames(cor_matrix), value = TRUE)
cor_matrix_subset <- cor_matrix[debt_vars, gdp_vars]

# Convert the full correlation matrix to a format suitable for ggplot
melted_subset_cor <- melt(cor_matrix_subset)

# Identify diagonal values for labeling
diagonal_values <- data.frame(Var1 = debt_vars, Var2 = gdp_vars, value = diag(cor_matrix_subset))

# Create the heatmap
ggplot(melted_subset_cor, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(data = diagonal_values, aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Heatmap: Government Debt vs. GDP Growth",
       x = "GDP Growth",
       y = "Government Debt")
```
From the heatmap, we can see that the government debt is mostly inversely correlated with GDP growth: higher levels of government debt are associated with lower GDP growth rates. For example, in year 2019 and 2020, the correlation between government debt in 2019 and GDP growth in 2020 are -0.36 and -0.33. This indicates that countries with higher government debt tend to experience slower GDP growth, which aligns with the economic theory that high debt burdens would constrain economic growth.

However, not all years exhibit negative correlation: the correlation in year 2022 is positive and it is close to 0 in year 2021 (the value is -0.001, so it’s rounded to 0). The anomalies in these two years can be explained by the pandemic and the economic conditions after it. Due to pandemic, the government spending were implemented to recover the economy. Therefore, despite high government debt levels, more of it was for stimulating the economy and bringing it back to the normal level. Thus, it can increase GDP growth and lead to slightly positive value for correlation in year 2022.



## Graph 8: GDP growth to debt ratio for all years
```{r}
# Calculate GDP growth to debt ratio for each year
mutated_data <- data %>%
  mutate(
    Ratio_2018 = GDP_Growth_2018 / Government_Debt_2018,
    Ratio_2019 = GDP_Growth_2019 / Government_Debt_2019,
    Ratio_2020 = GDP_Growth_2020 / Government_Debt_2020,
    Ratio_2021 = GDP_Growth_2021 / Government_Debt_2021,
    Ratio_2022 = GDP_Growth_2022 / Government_Debt_2022,
    Ratio_2023 = GDP_Growth_2023 / Government_Debt_2023
  )

# Filter dataset for only European countries
european_data <- mutated_data %>%
  filter(Region == "Europe") # Adjust "Region" to match your column name

# Select relevant columns
european_selected_data <- european_data %>%
  select(Country, Ratio_2018, Ratio_2019, Ratio_2020, 
         Ratio_2021, Ratio_2022, Ratio_2023)

# Remove missing values
european_selected_data <- na.omit(european_selected_data)

# Create the parallel coordinates plot with the capped data
parcoords(
  european_selected_data,
  brushMode = "1D-axes",   # Enable brushing
  rownames = FALSE,        # Hide rownames
  reorderable = TRUE,      # Allow reordering of axes
  queue = TRUE,            # Enable smoother transitions
  alpha = 0.5              # Set transparency for overlapping lines
)
```
To further understand the relationship between government debt levels and economic growth, we focused on countries of Europe. The parallel coordinates plot visualizes the ratio of GDP growth to government debt for European countries across six years (2018–2023). 

From the plot, most countries maintain relatively stable ratios between 0 and 0.2, indicating that most countries balanced economic growth with their debt levels. However, in 2020, the ratio decreases for most European countries, where many countries even have negative values. This reflects the economic disruptions caused by the COVID-19 pandemic, which significantly reduced GDP growth for most countries though they increased government debt to help recover economy. In 2021, the economy bounced back, leading to many countries having ratio between 0.1 to 0.2, which indicated that more government debt spent on economic recovery.

There are certain countries consistently deviating from the general trends. For example, countries like Greece and Italy tend to have consistently lower ratios, suggesting slower economic growth for these countries. In 2020, extreme negative ratios for countries like Kosovo and Andorra indicates the disproportionate economic impact of the pandemic.


## Graph 9: Unemployment_Rate of developed and developing countries (2018-2023)
```{r}
filtered_data <- data %>%
  select(Status, starts_with("Unemployment_")) %>%
  pivot_longer(
    cols = starts_with("Unemployment_"),
    names_to = "Year",
    values_to = "Unemployment_Rate"
  ) %>%
  na.omit()

ggplot(filtered_data, aes(x = Status, y = Unemployment_Rate, fill = Status)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Unemployment Rate Comparison",
    x = "Country Status",
    y = "Unemployment Rate (%)"
  ) +
  scale_fill_manual(values = c("Developed" = "green", "Developing" = "pink")) +
  facet_wrap(~Year, ncol = 3) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```
The boxplots from 2018 to 2023 compare unemployment rates between developed and developing countries, revealing noticeable disparities. Developed countries consistently show lower median unemployment rates than developing countries each year. Furthermore, the narrower interquartile range (IQR) for developed nations indicates more stable unemployment rates, while the wider IQR for developing countries highlights greater variability.

Moreover, developing countries have more outliers. For example, in year 2023, one country reached more than 40% unemployment rate. In contrast, developed have signficantly less fewer outliers. In the same year 2023, one developed country reached almost 20% unemployment rate, which is just half of 40%. Therefore, these boxplots highlights the difference faced by developing and developed countries in maintaining employment stability.



## Graph 10: Inflation Rate of developed and developing countries in 2023
```{r}
# Average the inflation
average_inflation_ <- data %>%
  rowwise() %>%
  mutate(
    Average_Inflation = mean(
      c_across(starts_with("Inflation_Rate_")), na.rm = TRUE
    )
  ) %>%
  ungroup()

# Bin inflation rates into specified categories and calculate proportions
inflation_binned <- average_inflation_ %>%
  select(Status, Average_Inflation) %>%
  mutate(
    Inflation_Bin = cut(
      Average_Inflation,
      breaks = c(0, 5, 10, 15, 20, Inf),
      labels = c("0-5", "5-10", "10-15", "15-20", "20+"),
      right = FALSE
    )
  ) %>%
  group_by(Status, Inflation_Bin) %>%
  summarise(
    Count = n(),
    .groups = "drop"
  ) %>%
  group_by(Status) %>%
  mutate(
    Proportion = Count / sum(Count, na.rm = TRUE)  # Calculate proportions correctly for each status
  ) %>%
  ungroup() %>%
  complete(Status, Inflation_Bin, fill = list(Count = 0, Proportion = 0))  # Ensure all bins are represented

# Create the bar plot with proportions
ggplot(inflation_binned, aes(x = Inflation_Bin, y = Proportion, fill = Status)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_fill_manual(
    values = c("Developed" = "green", "Developing" = "skyblue"),
    name = "Country Status"
  ) +
  theme_minimal() +
  labs(
    title = "Proportional Distribution of Average Inflation Rates (2018-2023)",
    x = "Inflation Rate Bin (%)",
    y = "Proportion of Countries"
  ) +
  scale_y_continuous(labels = scales::percent) +  # Format y-axis as percentages
  theme(
    legend.position = "right",                     # Move legend to the right
    plot.title = element_text(hjust = 0.5)         # Center align the title
  )

```
The graph shows that both developed and developing countries predominantly have average inflation rates in the "0-5%" and "5-10%" ranges, but their distributions tell different stories. For developed countries, over 60% fall within the "0-5%" range, reflecting their consistent ability to maintain low inflation. A smaller percentage of developed nations experience inflation in the "5-10%" range, and very few exceed 10%, emphasizing their economic stability.

On the other hand, while about 55% of developing countries also fall into the "0-5%" range, their inflation rates are much more spread out. Unlike developed countries, a significant number of developing nations have inflation rates in the "10-15%" and "20%+" ranges. This highlights the higher inflation pressures many developing countries face, with some even grappling with hyperinflation, as shown by the "20%+" category. This disparity underscores the economic challenges that developing nations encounter compared to their more stable counterparts.
